Classical Dynamics of Harmonically Trapped Interacting Particles

### 2017-12-15 draft_07.tex
Reasonably complete draft

\documentclass[onecolumn,pra]{revtex4-1}

@@

\bibliographystyle{apsrev4-1}

\usepackage{amsmath}
 \usepackage{amsfonts}
 \usepackage{amssymb}
 \usepackage{graphicx}
 \usepackage{subfigure}
 \usepackage{enumerate}
 \usepackage{verbatim}% for \begin{comment} \end{comment}
 \usepackage{color}
 \usepackage{xcolor}
 \usepackage{bm}
 \usepackage[]{hyperref}
 \hypersetup{
    colorlinks = false,
    linkbordercolor = {red},
    citebordercolor = {green},
}

@@

\begin{document}
\title{One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction}

@@

Abstract --  to be written last

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation (CITE).  There are
also fundamental reasons for studying dynamics in isolation: recent activity has led to advances in
the understanding of basic principles connecting quantum dynamics, quantum chaos and the emergence
of statistical mechanics and thermalization (CITE).  For classical many-body systems, explorations
of the connections between microscopic dynamical rules, chaos, ergodicity and statistical mechanics
have a much longer history, dating back to the time of Boltzmann and Poincare.
\cite{Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85}.

Motivated by questions and issues arising in the field of quantum many-body and few-body dynamics,
in this work we will study the classical dynamics of a collection of interacting particles confined
in a one-dimensional harmonic trap.  The analogs of two questions, studied intensively in the
quantum literature, will be investigated: (1) the influence of interactions on the breathing mode
frequency of the trapped gas, and (2) the relaxation and possible thermalization of the finite
system under its own evolution, in the absence of any external baths or dissipation mechanisms.

Experiments with cold atoms, which have had an essential motivational role for the current interest
in many-body quantum dynamics, are generally performed in a trapping potential, either optical or
magnetic.  As a result, in the last two decades many questions of many-body quantum physics have
been reformulated in the presence of a harmonic trap.  The presence of a harmonic trap often results
in new dynamical phenomena; for example the motion of vortices in a trapped condensate (CITE) has
new features (such as precession) compared to vortex motion in uniform condensates (CITE).  Trapping
also results in new collective phenomena which have no analog in uniform situations: for example,
collective modes like dipole, breathing and quadrupole modes are specific to trapped many-body
systems.  Dipole modes (center of mass oscillations) and breathing modes (size oscillations) appear
even for one-dimensional trapped systems; higher-dimensional systems display additional, more
complicated modes like quadrupole and scissors modes.  Because of the omnipresence of trapping
potentials in cold-atom experiments, dipole and breathing mode dynamics are pervasive across the
field: These modes have been intensively studied and used for diagnostic purposes since the early
days of research with trapped quantum-degenerate gases, and continue to generate interest today
\cite{Naegerl_Science2009, Esslinger_exp_PRL2003, collective_modes_early_papers,
  Tschischik_BHbreathing_PRA_2013, KroenkeSchmelcher_BM, Bonitz_Review, Bouchoule_PRL2014,
  our_superTG_PRA_2015, STG_Astrakharchik_PRL2005, STG_Chen-etal_PRA2010,
  br_mode_long_range_interactions, 1D_breathing_mode_recent, KohnMode, Cominotti_2015,
  ZarembaTowers_PRA2014, Wetterich_JPB2011, fermion2Dbreathing, Stringari_lowDcollective_PRA15,
  Mazets_1Dbreathing_EPJD11}.  (CITE).
%
In one of the best-known experiments on quantum many-body dynamics,
Ref.~\cite{NewtonsCradle_Nature06}, the dynamics of an integrable system (Lieb-Liniger bosons) was
seen to have ultraslow relaxation in a trap, supposedly due to integrability.  This experiment has
led to continuing efforts to understand how a trap breaks integrability (CITE).

The presence of harmonic trapping is thus a paradigmatic aspect in the field of quantum many-body
dynamics.  This motivates our study of classical dynamics of trapped interacting systems.  Similarly
motivated by quantum dynamics, classical trapped dynamics of hard rods has recently been studied in
Refs.\ \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10}, focusing on integrability-breaking effects of
the trap.  In this work, we consider trapped classical particles which interact via a simple
finite-range potential between any pair of particles: a constant repulsive force acts whenever the
particles are less than a certain distance apart.  The interaction strength can be tuned, leading to
different regimes of behavior.

Our first theme will be a study of the breathing mode of the classical interacting gas, and the
effect of interactions on the frequency of the breathing mode.  In the quantum case, this has long
been a topic of interest for both fermions and bosons.  For bosons with short-range interactions,
the situation has been studied in one, two and three dimensions.  At zero interactions, the
breathing mode is twice the trapping frequency $\omega_0$ (assuming isotropic trapping) in each
case.  A mean field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the
trapping frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the
mean-field description can be expected to be qualitatively correct, thus the breathing mode
frequency is expected to change monotonically from $2\omega_0$ at zero interactions to
$\sqrt{5}\omega_0$ at large interactions.  For 2D traps, the breathing frequency is exactly
$2\omega_0$ for any interaction strength \cite{PitaevskiiRosch_PRA97}.  This remarkable result
follows from a symmetry in 2D.  Classical trapped particles with scale-invariant interaction
potentials also have breathing frequency $2\omega_0$ independent of interaction strength
\cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped Lieb-Liniger gas) is also
remarkable: as the interaction is increased, the breathing frequency first decreases from
$2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger interactions
increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau) limit
\cite{Naegerl_Science2009, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{br_mode_long_range_interactions, Bonitz_Review}.  The breathing mode has also
been studied for a high-temperature trapped gas where a Boltzmann-equation description is valid,
both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

In this work, we use an explicit microscopic model, rather than a Boltzmann equation approach.
Unlike the quantum case, the harmonic potential does not impose a length scale or an energy scale.
The interaction range can thus be scaled away by a redefinition of length, i.e., by measuring
lengths in units of the interaction range.  There are only two important parameters determining the
system behavior, namely the interaction strength and the energy per particle.  We find that the
breathing mode frequency has a non-monotonic dependence on the interaction parameter.  At small
interactions, when the particles pass through each other, the effect of collisions is to slow down
the dynamics, so that the breathing frequency drops below the non-interacting value $2\omega_0$.  At
very large interactions, the particles bounce off each other, a process which speeds up the size
oscillations, resulting in the breathing frequency being larger than $2\omega_0$.

Our second theme is the process of relaxation at long timescales, and possible thermalization.  This
topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represents the thermal value (CITE).
The connection to statistical mechanics is particularly difficult for explicitly finite systems, in
which case ETH is to be understood in terms of finite size scaling (CITE).  In classical systems,
the connection between microscopic dynamics and statistical mechanics has been continually studied
since the 19th century.  It is generally understood that systems with positive Lyapunov exponents

---------

Stat-mech from the `bottom up'.  Find papers that do the same.

More detailed citations of breathing mode.  There are also papers on classical breathing mode, e.g.,
by Bonitz.

Controlled and tunable experimental realizations of confined quantum systems with ultracold atoms
enable nonequilibrium studies of quantum many-body states in novel geometries. A common feature of
many cold-atom experiments is the confinement of a many-body system in a harmonic trap. Trapping
introduces many distinctive features which have no analog in uniform many-body states, including
collective excitations such as breathing modes, dipole modes, and scissors modes. Such trap-related
collective modes have been widely studied both experimentally and theoretically for continuum
systems, especially in the mean-field regime, since the early days after quantum degeneracy was
achieved with trapped atoms \cite{Dalfovo1997}\cite{Jin1996}.

In mean-field perspective, the physics of atom gas in Bose-Einstein Condensation can be described by
the Gross-Pitaevskii equation\cite{Dalfovo1999}\cite{Stringari1996}, where the breathing frequency
for 1D Bose gas is calculated to be $\Omega_{GP}=\sqrt{3}\omega_0$. Meanwhile, the experiment of
Ref.\cite{Haller2009} has found a regime of interactions where the breathing-mode frequency
approaches the Gross-Pitaevskii prediction. However, since mean-field approach which generally omit
quantum fluctuation is successful in this problem, one may wonder how quantum effects play the role
in the breathing mode. Thus, a natural question would be asked is that how breathing frequency is
dependent on interaction for classical gas. Though generated in BEC, the breathing mode phenomenon
is not restricted to BEC. For a classical gas, the breathing mode still can be excited with a
quench. Usually, a good knowledge of the classical version may help us understand the quantum one
better. David Guery-Odelin, Francesca Zambelli, Jean Dalibard, and Sandro Stringari answered this
question in ref.\cite{Guery-Odelin1999}. They used the Boltzmann equation as well as molecular
dynamics simulation to study the monople mode and quarulple mode of the 2-D classical gas in a
harmonic trap. But in that work, the interaction is elastic collision,
i.e. zero-range-infinite-strength interaction. We think introducing a finite range for the repulsive
interaction will definitely enrich the physics in this system which looks simple, thus producing a
different equilibrium and non-equilibrium behaviour. Surprisingly, to the best of our knowledge,
there is no in-depth study of this case. For this reason, more study is necessary.

On the other hand, in a nonquilibrium many-body system, a more intriguing question is about its
thermalization or relaxation property. Thermalization is among the most fundamental processes for a
many-body interacting system. All systems do not thermalize. An example is the Fermi-Pasta-Ulam
paradox [ ] which shows confinement in phase-space so that the system stay non-thermalized for a
long time. A lot of research on thermalization property has been done in other classical
system. Ref.\cite{Tsuchiya2000}\cite{Yawn1997} studied the one dimensional gravitational system with
Lyapunov exponents. Ref.\cite{Jin2013} studied the ring of harmonic oscillators and magnetic moment
by examine the canonical distribution of a subsystem . In our paper, we will also probe the system's
thermalization time scales from the perspective of canonical distribution as well as Lyapunov
exponent.

%With these approaches, we will show that the thermalization time scale manifest a
%competition between total energy and interaction energy, so there is not any singular
%behavior in the thermalization time scales except the trivial divergence at zero
%interaction.

------------------

Breathing frequency, comparisong to quantum case ------

Usually, to see the response of quantum gas to a quench, one begin simulation with the ground state
and then suddenly vary some parameter by a little, for instance,
$\omega_0'\rightarrow\omega_0=\omega_0'+\Delta\omega$. In that case, one will observe the radius R
of the cloud oscillating at a certain frequency, which is usually called ``breathing mode".

However, it is quite different in the classical version. Firstly, it is no longer necessary to focus
on ground state. In quantum case, the reason why we care about ground state and low lying
excitations is that BEC occur at low temperature. But in classical case, there is no BEC. So it is
not necessary to focus on low lying excitations in classical case. Besides, since classical limit is
$\hbar\rightarrow 0$, no matter how low the kT is, $\hbar/kT$ is always zero so even in low energy
regime, the classical behaviors could be quite different from the quantum one. Secondly, it is no
longer necessary to vary parameters ``by a little" in the quench. In quantum case, the small quench
$\omega_0'\rightarrow\omega_0=\omega_0'+\Delta\omega$ means the spectrum of Hamiltonian is shifted
slightly, thus one may expect to see some beat phenomenon({\color{red}{?}}). But in classical case,
since $\hbar$ vanishes, no matter how small the quench is, the reason that cause beats in quantum
case no longer work in classical context. Thus it is not necessary to limit ourselves to small
quench.

Let's consider quenching the harmonic potential, i.e.,
$\omega_0\rightarrow\omega_0'=\omega_0+\Delta\omega_0$. We notice that X and P (scaled) are
symmetric for simple harmonic oscillator. In weak interaction limit, the distribution cloud should
be circular when the gas is completely thermalized. The difference between thermalized cloud before
and after quench is that they are circular under different scaling ratio of P and X axis. If we look
at both of them in the phase-space rescaled according to $\omega_0'$, we will find shape of the
cloud before the quench is an ellipse while the final one is a circle.

------------------

The words ergodic, relaxation and thermalization

Loss of memory of initial state --- our two definitions

------------------

This paper is organized as follows: In Section \ref{section:preparation}, we perform a scaling to
reduce the number of parameters of the system. In addition, we briefly observe the ground state and
reveal two dynamically distinct regime of the system, i.e. strong interaction regime and weak
interaction regime. In Section \ref{section:breathing frequency}, we evaluate the breathing
frequency according to its mechanism, and compare that with our numerical simulation
result. Besides, we also introduce a rotating frame method of treating this problem, which is a
convenient tool to analyze simple harmonic systems. In section \ref{section:Thermalization}, we
focus on the thermalization property. We first examine the canonical distribution of energy to find
out the threshold of thermalization, i.e. under what parameter the thermalization time scale is very
large. In the end of this section, we measure the Lyapunov Exponent to quantify the time scales and
confirm the former discussion. Then we point out that there are two kind of thermalization behavior
that could be measured in this system, one is about energy distribution while another is about shape
of cloud in phase-space. Since here we find two sense of ``thermalizing", we will redefine our terms
-- When a system's energy distribution goes to equilibrium, we still call it ``thermalized".  But
when a system's distribution in phase-space goes to equilibrium, we call it ``relaxed" to distiguish
it from ``thermalized". So in section \ref{section:Shape} we actually study the ``thermalizing" of
the shape of the cloud in phase-space, or the ``relaxation" behavior of the system. There we
discovered a long-lasting mode, which turns out to be a quadruple mode in phase-space.

\section{The model and its equilibrium properties}\label{section:preparation}

@@

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
H=\frac{1}{2}m\omega_0^2\sum_{i}x_i^2+\frac{1}{2}m\sum_i
v_i^2+\sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
  \tilde{H} =
  \frac{1}{2}\sum_{i}\tilde{x_i}^2
  + \frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2 
  + \sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless
    1}\tilde{F}_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)  
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``E" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R\equiv\left(\overline{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.

In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place.  It is also expected to be a good approximation at very
small $F_0$ since we can then simply neglect the interaction energy.  Thus, we will use this as a
common estimate for $R$ in terms of the energy.

@@

\subsection{Phase space snapshots}

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.pdf}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

>>>>

\documentclass[a4paper, onecolumn]{revtex4-1}

@@

\usepackage{amsmath}
 \usepackage{amsfonts}
 \usepackage{amssymb}
 \usepackage{graphicx}
 \usepackage{subfigure}
% \usepackage{enumerate}

% \usepackage{verbatim}% for \begin{comment} \end{comment}

\usepackage{color}
\usepackage{xcolor}
\usepackage{bm}
\usepackage[]{hyperref}
\hypersetup{
    colorlinks = false,
    linkbordercolor = {red},
    citebordercolor = {green},
}

@@

\begin{document}

\title{Classical Dynamics of Harmonically Trapped Interacting Particles}

@@

  Motivated by the interest in the dynamics of trapped quantum gases, we study the microcanonical
  dynamics of a trapped gas of classical particles interacting with a finite-range interaction of
  tunable strength.  We examine two questions which have been of interest in quantum dynamics: (1)
  the breathing mode (size oscillation) dynamics of the trapped gas and the dependence of the
  breathing frequency on the interaction strength, and (2) the long-time relaxation and possible
  thermalization of the finite isolated gas.

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation.  There are also
foundational reasons for studying quantum dynamics in isolation: recent activity has led to advances
in the understanding of basic principles connecting quantum dynamics, quantum chaos and the
emergence of statistical mechanics and thermalization (CITE).  For classical many-body systems,
explorations of the connections between microscopic dynamical rules, chaos, ergodicity and
statistical mechanics have a much longer history, dating back to the time of Boltzmann and Poincare
\cite{Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85}.

Motivated by questions and issues arising in the field of quantum many-body and few-body dynamics,
in this work we will study the classical dynamics of a collection of interacting particles confined
in a one-dimensional harmonic trap.  The analogs of two questions, studied intensively in the
quantum literature, will be investigated for this classical system: (1) the influence of
interactions on the breathing mode frequency of the trapped gas, and (2) the relaxation and possible
thermalization of the finite system under its own evolution, in the absence of any external baths or
dissipation mechanisms.

Experiments with cold atoms, which have an essential motivational role for the current interest in
many-body quantum dynamics, are generally performed in a trapping potential, either optical or
magnetic.  As a result, in the last two decades many questions of many-body quantum physics have
been reformulated in the presence of a harmonic trap.  The presence of a harmonic trap often results
in new dynamical phenomena.  For example, the motion of vortices in a trapped condensate (CITE) has
new features, such as precession, compared to vortex motion in uniform condensates (CITE).  Trapping
also results in new collective phenomena which have no analog in uniform situations: for example,
collective modes like dipole, breathing and quadrupole modes are specific to trapped many-body
systems.  Dipole modes (center of mass oscillations) and breathing modes (size oscillations) appear
even for one-dimensional trapped systems; higher-dimensional systems display additional, more
complicated modes like quadrupole and scissors modes.  Because of the omnipresence of trapping
potentials in cold-atom experiments, dipole and breathing mode dynamics are pervasive across the
field: These modes have been intensively studied and used for diagnostic purposes since the early
days of research with trapped quantum-degenerate gases, and continue to generate interest today
\cite{Naegerl_Science2009, Esslinger_exp_PRL2003, collective_modes_early_papers,
  Tschischik_BHbreathing_PRA_2013, KroenkeSchmelcher_BM, Bonitz_Review, Bouchoule_PRL2014,
  our_superTG_PRA_2015, STG_Astrakharchik_PRL2005, STG_Chen-etal_PRA2010,
  br_mode_long_range_interactions, 1D_breathing_mode_recent, KohnMode, Cominotti_2015,
  ZarembaTowers_PRA2014, Wetterich_JPB2011, fermion2Dbreathing, Stringari_lowDcollective_PRA15,
  Mazets_1Dbreathing_EPJD11}.  (CITE).
%
In one of the best-known experiments on quantum many-body dynamics,
Ref.~\cite{NewtonsCradle_Nature06}, the dynamics of an integrable system (Lieb-Liniger bosons) was
seen to have ultraslow relaxation in a trap, supposedly due to integrability.  This experiment has
led to continuing efforts to understand how a trap breaks integrability (CITE).

The presence of harmonic trapping is thus a paradigmatic aspect in the field of quantum
non-equilibrium physics.  This motivates our study of classical dynamics of trapped interacting
particles.  Similarly motivated by quantum dynamics, classical trapped dynamics of hard rods has
recently been studied in Refs.\ \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10}, focusing on
integrability-breaking effects of the trap.  In this work, we consider trapped classical particles
which interact via a simple finite-range potential between any pair of particles: a constant
repulsive force acts whenever the particles are less than a certain distance apart.  The interaction
strength can be tuned, leading to different regimes of behavior.

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interactions, the
breathing mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.
A mean field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the
trapping frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the
mean-field description can be expected to be qualitatively correct, thus the breathing mode
frequency is expected to change monotonically from $2\omega_0$ at zero interactions to
$\sqrt{5}\omega_0$ at large interactions.  For 2D traps, the breathing frequency is exactly
$2\omega_0$ for any interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical
trapped particles with scale-invariant interaction potentials also have breathing frequency
$2\omega_0$ independent of interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction
dependence in 1D (trapped Lieb-Liniger gas) is also remarkable: as the interaction is increased, the
breathing frequency first decreases from $2\omega_0$ to the mean-field prediction
$\sqrt{3}\omega_0$, and then at larger interactions increases again, returning to $2\omega_0$ in the
infinite interaction (Tonks-Girardeau) limit \cite{Naegerl_Science2009,
  Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent, Bouchoule_PRL2014,
  KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{br_mode_long_range_interactions, Bonitz_Review}.  The breathing mode has also
been studied for a high-temperature trapped gas where a Boltzmann-equation description is valid,
both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

In this work, we use an explicit microscopic model treated microcanonically, rather than a Boltzmann
equation approach.  Unlike the quantum case, the harmonic potential does not impose a length scale
or an energy scale.  The interaction range can thus be scaled away by a redefinition of length,
i.e., by measuring lengths in units of the interaction range.  There are thus only two important
parameters determining the system behavior, namely the interaction strength and the energy per
particle.  We find that the breathing mode frequency has a non-monotonic dependence on the
interaction parameter.  At small interactions, when the particles pass through each other, the
effect of collisions is to slow down the dynamics, so that the breathing frequency drops below the
non-interacting value $2\omega_0$.  At very large interactions, the particles bounce off each other,
a process which speeds up the size oscillations, resulting in the breathing frequency being larger
than $2\omega_0$.

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represents the thermal value (CITE).
The connection to statistical mechanics is particularly difficult for explicitly finite systems, in
which case ETH is to be understood in terms of finite size scaling (CITE).  In classical systems,
the connection between microscopic dynamics and statistical mechanics has been continually studied
since the 19th century \cite{Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book,
  EckmannRuelle_RMP85}.  It is generally understood that systems with nonzero Lyapunov exponents,
i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided there are enough
degrees of freedom.  As in the quantum case, finite systems are particularly intricate.  In finite
systems, defining thermalization is trickier as most observables will fluctuate or oscillate
substantially.  In addition, the influence of Kolmogorov-Arnold-Moser (KAM) orbits may play a role
in preventing thermalization; it is generally believed that the fraction of phase space where KAM
physics might be relevant decreases rapidly with particle number \cite{Dumas_book_KAMstory}.

In our work, we examine the relaxation of our trapped interacting classical system, explicity for
finite numbers of particles.  We look for thermalization within finite timescales, by sampling
single-particle energies periodically until such timescales and comparing the distribution of
single-particle energies with the Boltzmann distribution.  We show that whether or not the system
thermalizes in this sense depends on the ratio of interaction strength and energy per particle.  We
also show that this result is reflected in the finite-time Lyapunov exponents of the system.
%
We also examine another notion of thermalization, according to which a system is thermalized when it
loses memory of initial state, by following the dynamics of the particle distribution in phase
space.  Remarkably, we find that one can observe energy thermaliztion even at timescales when the
shape of the phase space distribution continues to perform seemingly coherent oscillations.

There is a significant literature on calculating Lyapunov exponents in model statistical-mechanics
systems and thus exploring ergodicity \cite{Dorfman_book_1999} (CITE).  This provides intuition of
what types of microscopic interactions are likely to produce thermalizability, and is an important
step in the program of understanding statistical mechanics ``from the bottom up''.  However, we are
not aware of a significant literature on following explicitly the microcanonical dynamics of
statistical-mechanical or few-body systems.  Several such studies have appeared recently (CITE),
motivated by the quantum dynamics literature, like the present work.  A few earlier studies have
also examined thermalization dynamics in one dimensional gravitational systems \cite{Yawn1997,
  Tsuchiya2000}.
%
Investigating such real-time dynamics can be expected to provide insights on the timescales and
microscopic mechanisms involved in the emergence of statistical mechanics from collections of
particles.

This paper is organized as follows: In Section \ref{sec:preparation}, we introduce the Hamiltonian,
identify and introduce the essential scales, and describe some aspects of the model and our
simulations.  The breathing frequency is treated in Section \ref{sec:breathing}.  We explain the
main features using both real-space and phase-space pictures of the interaction process, obtain
estimates for the interaction- and energy-dependence of the frequency shift, and compare these
predictions with numerics.  Section \ref{sec:relaxation} treats relaxation and thermalization.  We
propose a condition for relaxation in reasonable timescales from considerations of few-body
dynamics, compare the energy distribution with the Boltzmann distribution, show how the largest
finite-time Lyapunov exponent distribution reflects the relaxation condition, and examine shape
dynamics of the particle distribution in phase space.  Concluding remarks appear in Section
\ref{sec:concl}.

%% This paper is organized as follows: In Section \ref{section:preparation}, we perform a scaling to
%% reduce the number of parameters of the system. In addition, we briefly observe the ground state and
%% reveal two dynamically distinct regime of the system, i.e. strong interaction regime and weak
%% interaction regime. In Section \ref{section:breathing frequency}, we evaluate the breathing
%% frequency according to its mechanism, and compare that with our numerical simulation
%% result. Besides, we also introduce a rotating frame method of treating this problem, which is a
%% convenient tool to analyze simple harmonic systems. In section \ref{section:Thermalization}, we
%% focus on the thermalization property. We first examine the canonical distribution of energy to find
%% out the threshold of thermalization, i.e. under what parameter the thermalization time scale is very
%% large. In the end of this section, we measure the Lyapunov Exponent to quantify the time scales and
%% confirm the former discussion. Then we point out that there are two kind of thermalization behavior
%% that could be measured in this system, one is about energy distribution while another is about shape
%% of cloud in phase-space. Since here we find two sense of ``thermalizing", we will redefine our terms
%% -- When a system's energy distribution goes to equilibrium, we still call it ``thermalized".  But
%% when a system's distribution in phase-space goes to equilibrium, we call it ``relaxed" to distiguish
%% it from ``thermalized". So in section \ref{section:Shape} we actually study the ``thermalizing" of
%% the shape of the cloud in phase-space, or the ``relaxation" behavior of the system. There we
%% discovered a long-lasting mode, which turns out to be a quadruple mode in phase-space.

\section{The model and its equilibrium properties}\label{sec:preparation}

@@

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
  H ~=~
  \frac{1}{2}m\omega_0^2\sum_{i}x_i^2
  ~+~ \frac{1}{2}m\sum_i  v_i^2
  ~+~ \sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega_0$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
  \tilde{H} =
  \frac{1}{2}\sum_{i}\tilde{x_i}^2
  + \frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2 
  + \sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless
    1}\tilde{F}_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)  
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``$E$" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R\equiv\left(\overline{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.
%
In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place, provided the gas is dilute.  It is also expected to be a
good approximation at very small $F_0$ since we can then simply neglect the interaction energy.
Thus, we will use this as a common estimate for $R$ in terms of the energy.

@@

\subsection{Phase space distribution}

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[tbph]
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.pdf}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%
